---
title: "L03 ggplot 2"
subtitle: "Data Visualization (STAT 302)" 
author: "Sherry Chen"
pagetitle: "L03 Chen Sherry"

format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    link-external-newwindow: true

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin
---



::: {.callout-warning}
This is a template file and may not contain everything seen in the instructions html (`L03_ggplot2.html`). For example, images or graphs are not usually included in the template. **Rely on the instructions html file for directions, not the template file.**
:::

## Overview

The goal of this lab is to continue the process of unlocking the power of `ggplot2` through constructing and experimenting with a few basic plots.

## Datasets

We will be using the BRFSS survey which was introduced in the last lab. The data was supplied in `cdc.txt` file and should be in your `/data` subdirectory. If not, you need to download the last lab to get this data file. As a reminder, the dataset contains 20,000 complete observations/records of 9 variables/fields, described below.

* `genhlth` - How would you rate your general health? (excellent, very good, good, fair, poor)
* `exerany` - Have you exercised in the past month? (`1 = yes`, `0 = no`) 
* `hlthplan` - Do you have some form of health coverage? (`1 = yes`, `0 = no`) 
* `smoke100` - Have you smoked at least 100 cigarettes in your life time? (`1 = yes`, `0 = no`) 
* `height` -  height in inches
* `weight` - weight in pounds
* `wtdesire` - weight desired in pounds
* `age` - in years
* `gender` - `m` for males and `f` for females

```{r}
#| label: load-packages

# Load package(s)
library(tidyverse)
library(readr)
library(ggplot2)
# Load data

cdc <- read.table("data/cdc.txt", sep = "|", header = TRUE)


```

## Exercise 1

Using the `cdc` dataset, we want to look at the relationship between height and weight. Recreate the following graphics as precisely as possible.

### Plot 1

*Hints:*

- Transparency is 0.2
- Minimal theme

::: {.callout-tip icon="false"}
## Solution

```{r}
ggplot(cdc, aes(x = height, y = weight, color = gender)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = 'lm')+
  theme_minimal() +
  labs(x = "Height(in)", y = "Weight(lbs)", title = "Relationship between height and weight")
```


:::

### Plot 2

*Hints:*

- linewidth = 0.7

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: Plot 2 
ggplot(cdc, aes(x = height, y = weight, color = gender)) +
  geom_density_2d(size = 0.7) +  # Creates density contour 
  theme_minimal() +
  labs(x = "Height(in)", y = "Weight(lbs)", title = "Relationship between height and weight")
  
```


:::

### Plot 3

*Hints:*

- `bins` set to 35

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: Plot 3

ggplot(cdc, aes(x = height, y = weight)) + 
  geom_hex(bins  = 35) +
  theme_minimal() +
  labs(x = "Height(in)", y = "Weight(lbs)", title = "Relationship between height and weight")
```


:::

### Plot 4

*Hints:*

- use a stat layer, not a geom layer
- `geom = "polygon"` 

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: Plot 4 

ggplot(cdc, aes(x = height, y = weight)) + 
  stat_density_2d(geom = "polygon", mapping = aes(fill = after_stat(level))) + 
  facet_wrap(~gender) +
  theme_minimal()
 
  
```


:::

## Exercise 2

Using the `cdc_means` dataset derived from the `cdc` dataset, recreate the following graphic as precisely as possible.

*Hints:*

- Hex color code `#56B4E9`
- 95% confidence intervals (1.96 or `qnorm(0.975)`)
- Some useful values: 0.1, 0.7


```{r} 
#| label: ex-2-plot

# data wrangling
# calc mean and se for CI
cdc_means <- cdc |>
  mutate(wtloss = weight - wtdesire) |>
  group_by(genhlth) |>
  summarize(
    mean = mean(wtloss),
    se = sd(wtloss) / sqrt(n())
  ) |>
  mutate(genhlth = fct_reorder(factor(genhlth), desc(mean)))
```


::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: ex-2-plot-with error bar 
ggplot(cdc_means, aes(x = genhlth, y = mean)) +
   geom_col(fill = "#56b4e9", width = 0.7) +
   geom_errorbar(aes(ymin = mean - (1.96 * se), ymax = mean + (1.96 * se)), 
                 position = position_dodge(1.96), 
                 width = 0.1) +
   labs(title = "Mean Desired Weight Loss by General Health",
        x = "General Health",
        y = "Mean Desired Weight Loss in lbs") +
   theme_minimal() +
   theme(axis.text.x = element_text(angle = 0, hjust = 1))
```


:::

## Exercise 3

Using the `cdc_weight_95ci` dataset derived from the `cdc` dataset, recreate the following graphic as precisely as possible.

*Hints:*

- Useful values: 0.1, 0.5
- Need to know CI formula

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: ex-3-plot

# data wrangling
# calculate mean, se, and margin of error for CI formula
cdc_weight_95ci <- cdc |>
  group_by(genhlth, gender) |>
  summarise(
    mean_wt = mean(weight),
    se = sd(weight) / sqrt(n()),
    moe = qt(0.975, n() - 1) * se
  ) |> 
  ungroup() |> 
   mutate(genhlth = factor(genhlth, levels = c("excellent", "very good", "good", "fair", "poor")))

position_jitterdodge <- position_jitterdodge(jitter.width = 0, dodge.width = 0.5)

# Create the plot
ggplot(cdc_weight_95ci, aes(x = mean_wt, y = gender, color = genhlth)) +
  geom_point(position = position_jitterdodge) +
  geom_errorbarh(aes(xmin = mean_wt - moe, xmax = mean_wt + moe), height = 0.1,position = position_jitterdodge) +
  theme_minimal() +
  ggtitle("General Health (self-reported)")



```

:::
